---
title: "Daily Load Forecasting Competition — July 2011"
subtitle: "ENV 797: Time Series Analysis for Energy and Environment Applications"
author: "Mingjie Wei, Abrao Pereira"
date: "April 24, 2026"
output:
  pdf_document:
    number_sections: true
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4
)

suppressPackageStartupMessages({
  library(readxl)
  library(dplyr)
  library(tidyr)
  library(lubridate)
  library(ggplot2)
  library(forecast)
  library(knitr)
  library(kableExtra)
})
```

# GitHub Repository

**Repository link:** [Wei_Pereira_ENV797_TSA_ForecastCompetition_S26](https://github.com/mingjie-wei/Wei_Pereira_ENV797_TSA_ForecastCompetition_S26/)

# Problem Statement

## Objective

This project aims to forecast daily residential electricity demand for July 2011 using historical behind-the-meter load data. The task is framed as a time series forecasting competition in which past demand patterns, as well as temperature and relative humidity, may improve predictive accuracy.

## Data Context

Three datasets are provided for the competition:

-   **Load data:** hourly electricity demand (Wh) for a single residential meter from January 2005 to June 2011.
-   **Temperature data:** hourly temperature readings (°F) from nearby weather stations covering the same period.
-   **Humidity data:** hourly relative humidity readings (%) from nearby weather stations covering the same period.

The load series is the target variable. Temperature and humidity are optional external predictors.

## Forecasting Task

The hourly series were aggregated to daily averages before modeling. Multiple candidate models were trained, validated on July 2010 (hold-out), and compared. The best-performing model was then retrained on the full dataset (through June 2011) to generate forecasts for July 1–31, 2011.

# Data and Preprocessing

## Raw Data Description

```{r data-import}
hour_cols <- paste0("h", 1:24)

load_raw <- read_excel("data/load.xlsx") |>
  mutate(date = as.Date(date))

temp_raw <- read_excel("data/temperature.xlsx") |>
  mutate(date = as.Date(date))

hum_raw <- read_excel("data/relative_humidity.xlsx") |>
  mutate(date = as.Date(date))

cat("Load rows:", nrow(load_raw), "| Date range:",
    as.character(min(load_raw$date)), "to", as.character(max(load_raw$date)), "\n")
cat("Temp rows:", nrow(temp_raw), "\n")
cat("Humidity rows:", nrow(hum_raw), "\n")
```

## Data Wrangling

Each dataset is stored in wide format with one row per day and columns `h1`–`h24` (or per station). We compute daily averages by row-wise mean across all hourly / station columns.

```{r preprocessing}
# ---- Load: daily mean across 24 hours ----
daily_load <- load_raw |>
  mutate(load_day = rowMeans(across(all_of(hour_cols)), na.rm = TRUE)) |>
  select(date, load_day) |>
  arrange(date)

# ---- Temperature: daily mean across all weather stations ----
temp_station_cols <- setdiff(names(temp_raw), c("date", "hr"))
daily_temp <- temp_raw |>
  mutate(temp_mean = rowMeans(across(all_of(temp_station_cols)), na.rm = TRUE)) |>
  group_by(date) |>
  summarise(temp_day = mean(temp_mean, na.rm = TRUE), .groups = "drop")

# ---- Humidity: daily mean across all weather stations ----
hum_station_cols <- setdiff(names(hum_raw), c("date", "hr"))
daily_hum <- hum_raw |>
  mutate(hum_mean = rowMeans(across(all_of(hum_station_cols)), na.rm = TRUE)) |>
  group_by(date) |>
  summarise(hum_day = mean(hum_mean, na.rm = TRUE), .groups = "drop")

# ---- Join all three daily series ----
daily <- daily_load |>
  left_join(daily_temp, by = "date") |>
  left_join(daily_hum,  by = "date") |>
  arrange(date)

cat("Daily rows:", nrow(daily), "| Missing load:", sum(is.na(daily$load_day)), "\n")
```

We aggregated hourly load observations to daily averages by taking the mean of the 24 hourly values for each day. Temperature and humidity were similarly averaged across all available weather stations and across the 24 hours of each day.

## Construction of Time Series Objects

```{r ts-object}
# msts() captures both weekly (7) and annual (365.25) seasonality.
# Start: January 1, 2005 (day 1).
daily_ts <- msts(
  daily$load_day,
  seasonal.periods = c(7, 365.25),
  start = c(2005, 1)
)
```

We used `msts()` rather than `ts()` because daily electricity demand exhibits two overlapping seasonal patterns: a weekly cycle (weekday vs. weekend effect) and an annual cycle (summer cooling peaks, winter heating peaks).

## Exploratory Analysis

### Raw Time Series

```{r eda-raw-series, fig.cap="Daily average electricity demand (2005--2011). The shaded band marks the July 2010 validation window."}
ggplot(daily, aes(x = date, y = load_day)) +
  annotate("rect",
           xmin = as.Date("2010-07-01"), xmax = as.Date("2010-07-31"),
           ymin = -Inf, ymax = Inf, fill = "steelblue", alpha = 0.15) +
  geom_line(color = "grey30", linewidth = 0.35) +
  labs(x = NULL, y = "Daily avg load (Wh)",
       title = "Residential daily electricity demand") +
  theme_minimal()
```

### Annual Seasonal Pattern

```{r eda-monthly-boxplot, fig.cap="Distribution of daily load by calendar month, pooled across all years (2005--2011). Boxes span the interquartile range; whiskers extend to 1.5 IQR."}
daily |>
  mutate(month = lubridate::month(date, label = TRUE)) |>
  ggplot(aes(x = month, y = load_day)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6, outlier.size = 0.5) +
  labs(x = NULL, y = "Daily avg load (Wh)",
       title = "Annual seasonal pattern: load by month") +
  theme_minimal()
```

### Weekly Seasonal Pattern

```{r eda-dow-plot, fig.cap="Average daily load by day of week (Mon--Sun), pooled across all years."}
daily |>
  mutate(dow = lubridate::wday(date, label = TRUE, week_start = 1)) |>
  group_by(dow) |>
  summarise(mean_load = mean(load_day, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = dow, y = mean_load)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  labs(x = "Day of week", y = "Mean daily load (Wh)",
       title = "Weekly seasonal pattern: load by day of week") +
  theme_minimal()
```

### MSTL Decomposition

The `mstl()` function applies a Loess-based decomposition that simultaneously extracts weekly and annual seasonal components, a trend, and a remainder.

```{r eda-decomposition, fig.height=7, fig.cap="MSTL decomposition of the daily load series into trend, weekly seasonal, annual seasonal, and remainder components."}
autoplot(mstl(daily_ts)) +
  labs(title = "MSTL decomposition of daily load") +
  theme_minimal()
```

### Autocorrelation Structure (ACF and PACF)

The ACF and PACF are computed on the **training series** (through 2010-06-30) to avoid look-ahead leakage. Lags extend to 56 days (\~8 weeks) to reveal the weekly periodicity at multiples of lag 7.

```{r eda-acf, fig.cap="Sample ACF of daily load (training period 2005--2010-06-30, lags 1--56). Dashed lines mark the 95\\% confidence bounds under the null of no autocorrelation."}
y_tr_eda <- ts(
  daily$load_day[daily$date <= as.Date("2010-06-30")],
  frequency = 7
)
ggAcf(y_tr_eda, lag.max = 56) +
  labs(title = "ACF -- daily load (training set)") +
  theme_minimal()
```

```{r eda-pacf, fig.cap="Sample PACF of daily load (training period 2005--2010-06-30, lags 1--56)."}
ggPacf(y_tr_eda, lag.max = 56) +
  labs(title = "PACF -- daily load (training set)") +
  theme_minimal()
```

### Relationship with Temperature

```{r eda-weather-plot, fig.cap="Scatter plot of daily load vs. daily mean temperature (°F). The blue curve is a LOESS smoother."}
ggplot(daily, aes(x = temp_day, y = load_day)) +
  geom_point(alpha = 0.2, size = 0.7, color = "tomato") +
  geom_smooth(method = "loess", se = FALSE, color = "steelblue", linewidth = 0.8) +
  labs(x = "Daily mean temperature (°F)", y = "Daily avg load (Wh)",
       title = "Load vs. temperature") +
  theme_minimal()
```

### Summary of Key Patterns

-   **Trend:** The MSTL trend component reveals a broadly stable mean load over 2005–2011 with no strong long-run drift, though mild inter-year fluctuations are visible, likely reflecting weather variability across years rather than a structural change in consumption behavior.
-   **Annual seasonality:** Load peaks sharply in summer (June–August), consistent with air-conditioning demand, and shows a secondary elevation in winter (December–February) attributable to heating and reduced daylight hours. The monthly boxplot confirms that July is among the highest-load months, making it a challenging forecast target with elevated and variable demand.
-   **Weekly seasonality:** The day-of-week bar chart shows consistently higher load on weekdays (Monday–Friday) relative to weekends, reflecting reduced commercial and occupational activity on Saturdays and Sundays. This weekly cycle is the dominant short-run seasonal signal.
-   **Autocorrelation:** The ACF decays slowly and displays prominent spikes at multiples of lag 7 (7, 14, 21, ...), confirming strong weekly periodicity. The PACF shows significant partial autocorrelations through approximately lag 3–4 before dropping toward zero, suggesting a low-order autoregressive structure in the non-seasonal component — consistent with the ARIMA component selected within TBATS.
-   **Temperature relationship:** The scatter plot and LOESS smoother reveal a positive, roughly monotone relationship between daily mean temperature and daily load, with load rising steeply above approximately 60 °F. This pattern is consistent with cooling-driven electricity consumption dominating summer demand, and motivates the potential use of temperature as an external predictor in regression-based models.

# Experimental Design

## Training, Validation, and Final Forecasting Setup

We followed the competition structure when partitioning the data:

| Period           | Dates                    | Purpose                     |
|------------------|--------------------------|-----------------------------|
| Training         | 2005-01-01 to 2010-06-30 | Model fitting               |
| Validation       | 2010-07-01 to 2010-07-31 | Hold-out evaluation (MAPE)  |
| Final training   | 2005-01-01 to 2011-06-30 | Refit before final forecast |
| Forecast horizon | 2011-07-01 to 2011-07-31 | Kaggle submission           |

```{r train-validation-split}
train_end <- as.Date("2010-06-30")
val_start <- as.Date("2010-07-01")
val_end   <- as.Date("2010-07-31")

train2010  <- daily |> filter(date <= train_end)
val2010    <- daily |> filter(date >= val_start, date <= val_end)
train_full <- daily |> filter(date <= as.Date("2011-06-30"))

cat("Training rows:", nrow(train2010),
    "| Validation rows:", nrow(val2010),
    "| Full-training rows:", nrow(train_full), "\n")
```

## Model Development Strategy

We began with TBATS to capture both weekly and annual seasonality simultaneously, then explored additional approaches contributed by both team members. [Insert brief rationale for Models 2–5 as you develop them.]

## Evaluation Criteria

-   **Primary metric:** Mean Absolute Percentage Error (MAPE), consistent with the Kaggle leaderboard.
-   **Secondary metric:** Root Mean Squared Error (RMSE) for scale-sensitive comparison.

```{r metric-functions}
mape_pct <- function(actual, predicted) {
  ok <- actual != 0 & is.finite(actual) & is.finite(predicted)
  100 * mean(abs((actual[ok] - predicted[ok]) / actual[ok]))
}

rmse_val <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2, na.rm = TRUE))
}
```

Performance was assessed on the July 2010 hold-out window and cross-checked against Kaggle scores after each submission.

## Reproducibility and Workflow

-   **GitHub** was used for code version control and collaboration.
-   **Kaggle** was used to submit and compare forecast outputs.
-   Individual model scripts (`scripts/forecast_TBATS.R`, and others) are also available in the repository.
-   Only the top five models are retained in the final knitted report.

# Modeling Results: Top 5 Models

```{r model-scores-init}
model_scores <- data.frame(
  Model         = character(),
  Predictors    = character(),
  Val_MAPE      = numeric(),
  Val_RMSE      = numeric(),
  Kaggle_MAPE   = character(),
  stringsAsFactors = FALSE
)
```

## Model 1: TBATS (dual seasonality: weekly + annual)

### Motivation

TBATS (Trigonometric seasonality, Box-Cox transformation, ARMA errors, Trend, Seasonal components) is designed for time series with multiple seasonal periods. By using `msts()` with periods 7 and 365.25, we aim to capture both weekly and annual cycles simultaneously — a structure well-suited to daily residential load data.

### Specification

We fitted `tbats()` on an `msts(seasonal.periods = c(7, 365.25))` object. TBATS automatically selects the number of Fourier harmonics for each seasonal period as well as whether to apply a Box-Cox transformation.

```{r model-1, cache=TRUE}
y_tr_tbats <- msts(train2010$load_day, seasonal.periods = c(7, 365.25))

fit_tbats <- tbats(y_tr_tbats)

fc_tbats_val <- as.numeric(forecast(fit_tbats, h = nrow(val2010))$mean)

mape_tbats <- mape_pct(val2010$load_day, fc_tbats_val)
rmse_tbats <- rmse_val(val2010$load_day, fc_tbats_val)

cat(sprintf("TBATS  -- Val MAPE: %.2f%%  |  Val RMSE: %.1f\n", mape_tbats, rmse_tbats))

model_scores <- rbind(model_scores, data.frame(
  Model       = "TBATS (m = 7, 365.25)",
  Predictors  = "Load only",
  Val_MAPE    = round(mape_tbats, 2),
  Val_RMSE    = round(rmse_tbats, 1),
  Kaggle_MAPE = "0.0998"
))
```

```{r model-1-plot, fig.cap="Model 1: TBATS forecast vs. observed (July 2010 validation)."}
data.frame(
  date      = val2010$date,
  observed  = val2010$load_day,
  forecast  = fc_tbats_val
) |>
  pivot_longer(-date, names_to = "series", values_to = "load") |>
  ggplot(aes(x = date, y = load, color = series)) +
  geom_line() +
  scale_color_manual(values = c(observed = "grey30", forecast = "tomato")) +
  labs(x = NULL, y = "Daily avg load (Wh)", color = NULL,
       title = "Model 1: TBATS -- July 2010 validation") +
  theme_minimal()
```

### Validation Performance

TBATS achieved a validation MAPE of **`r round(mape_tbats, 2)`%** and RMSE of **`r round(rmse_tbats, 1)`** on the July 2010 hold-out.

### Strengths and Limitations

-   **Strengths:** Captures dual seasonality (weekly + annual) without requiring manual specification; handles Box-Cox transformation automatically.
-   **Limitations:** Slower to fit than simpler models; less interpretable parameter structure.

------------------------------------------------------------------------

## Model 2: Seasonal Naive (as baseline) 

### Motivation

Seasonal naive is used as part of our model because its simple and depends on the same time of the past model. We use daily data with and annual cycle, the forecast for July 1 2011 is observed from July 1 2010. Our result shows that despite it simplicity our Naive model shows a strong output and act as a benchmark.

### Specification

We applied \`snaive()\` from the \`forecast\` package to an
\`msts(seasonal.periods = c(7, 365.25))\` training object.
The model requires no parameter estimation.

```{r model-2, eval=FALSE}

y_tr_snaive   <- msts(train2010$load_day,
                      seasonal.periods = c(7, 365.25))
fit_snaive    <- snaive(y_tr_snaive, h = nrow(val2010))
fc_snaive_val <- as.numeric(fit_snaive$mean)

mape_snaive <- mape_pct(val2010$load_day, fc_snaive_val)
rmse_snaive <- rmse_val(val2010$load_day, fc_snaive_val)

cat(sprintf("Seasonal Naive -- Val MAPE: %.2f%%  |  Val RMSE: %.1f\n",
            mape_snaive, rmse_snaive))

model_scores <- rbind(model_scores, data.frame(
  Model       = "Seasonal Naive",
  Predictors  = "Load only",
  Val_MAPE    = round(mape_snaive, 2),
  Val_RMSE    = round(rmse_snaive, 1),
  Kaggle_MAPE = "—"
))

data.frame(
  date     = val2010$date,
  observed = val2010$load_day,
  forecast = fc_snaive_val
) |>
  pivot_longer(-date, names_to = "series", values_to = "load") |>
  ggplot(aes(x = date, y = load, color = series)) +
  geom_line() +
  scale_color_manual(values = c(observed = "lightgrey",
                                forecast = "blue")) +
  labs(x = NULL, y = "Daily avg load (Wh)", color = NULL,
       title = "Model 2: Seasonal Naive -- July 2010 validation") +
  theme_minimal()




```

### Validation Performance

Based on our Naive model results, it shows that we got a validation MAPE of 19.36% and RMSE of 950.4. This is the weakest result among all five models,
confirming that more sophisticated approaches are needed to
beat the baseline.

### Strengths and Limitations

Our seasonal Naive is simple and requires no parameter estimation, forecasting based on a single past value from one year ago. However, it cannot adapt to weather variations or year-to-year changes in load level, which explains its weakest performance among all five models.

------------------------------------------------------------------------

## Model 3: SARIMA

### Motivation

For our third model, we try to incorporate what we found on our benchmark model to handle the seasonal correlation by use SARIMA extends ARIMA. The
ACF plot showed strong spikes at multiples of lag 7, confirming a weekly seasonal pattern. Since auto.arima() cannot directly process msts() objects with multiple seasonal periods, we created a weekly ts() object (frequency = 7) to capture the dominant short-run pattern. Orders were selected automatically via AIC minimization.

### Specification

We used auto.arima() with stepwise = TRUE and approximation =
TRUE on a weekly ts object (frequency = 7). The function
automatically selected ARIMA(4,0,1)(2,0,0)[7].

```{r model-3, eval=FALSE}
y_tr_sarima <- ts(train2010$load_day, frequency = 7,
                  start = c(2005, 1))

fit_sarima    <- auto.arima(y_tr_sarima,
                            stepwise      = TRUE,
                            approximation = TRUE)
print(fit_sarima)
checkresiduals(fit_sarima)

fc_sarima_val <- as.numeric(
  forecast(fit_sarima, h = nrow(val2010))$mean)

mape_sarima <- mape_pct(val2010$load_day, fc_sarima_val)
rmse_sarima <- rmse_val(val2010$load_day, fc_sarima_val)

cat(sprintf("SARIMA -- Val MAPE: %.2f%%  |  Val RMSE: %.1f\n",
            mape_sarima, rmse_sarima))

model_scores <- rbind(model_scores, data.frame(
  Model       = "SARIMA(4,0,1)(2,0,0)[7]",
  Predictors  = "Load only",
  Val_MAPE    = round(mape_sarima, 2),
  Val_RMSE    = round(rmse_sarima, 1),
  Kaggle_MAPE = "0.1437"
))

data.frame(
  date     = val2010$date,
  observed = val2010$load_day,
  forecast = fc_sarima_val
) |>
  pivot_longer(-date, names_to = "series", values_to = "load") |>
  ggplot(aes(x = date, y = load, color = series)) +
  geom_line() +
  scale_color_manual(values = c(observed = "lightgrey",
                                forecast = "blue")) +
  labs(x = NULL, y = "Daily avg load (Wh)", color = NULL,
       title = "Model 3: SARIMA -- July 2010 validation") +
  theme_minimal()
```

### Validation Performance

Now, in this model, our value does improve much better. SARIMA achieved a validation MAPE of 13.52% and RMSE of 690.4, improving upon the seasonal naive baseline. The Ljung-Box test (p = 0.081) confirmed residuals approximate white noise, indicating the model adequately captured the weekly autocorrelation structure.

### Strengths and Limitations

Our SARIMA model performs well in capturing weekly autocorrelation structure with orders selected automatically via AIC. However, it is limited to a single seasonal period (frequency = 7) and does not explicitly model the annual cycle, which may explain its weaker performance in July a month dominated by annual seasonality.

------------------------------------------------------------------------

## Model 4: STL + ETS

### Motivation

We added this model to our list because removing both the weekly and
annual seasonal components simultaneously. An ETS exponential
smoothing model is then fitted to the seasonally adjusted series. This two stage approach handles multiple seasonality through STL while using ETS for the remaining trend and error structure.

### Specification

We applied stlf() directly to the msts training object with
seasonal.periods = c(7, 365.25). The function performs STL
decomposition and ETS fitting in a single call, automatically
selecting the best ETS model form.

```{r model-4, eval=FALSE}
y_tr_stlf   <- msts(train2010$load_day,
                    seasonal.periods = c(7, 365.25))
fit_stlf    <- stlf(y_tr_stlf, h = nrow(val2010))
fc_stlf_val <- as.numeric(fit_stlf$mean)

mape_stlf <- mape_pct(val2010$load_day, fc_stlf_val)
rmse_stlf <- rmse_val(val2010$load_day, fc_stlf_val)

cat(sprintf("STL+ETS -- Val MAPE: %.2f%%  |  Val RMSE: %.1f\n",
            mape_stlf, rmse_stlf))

model_scores <- rbind(model_scores, data.frame(
  Model       = "STL + ETS",
  Predictors  = "Load only",
  Val_MAPE    = round(mape_stlf, 2),
  Val_RMSE    = round(rmse_stlf, 1),
  Kaggle_MAPE = "—"
))

data.frame(
  date     = val2010$date,
  observed = val2010$load_day,
  forecast = fc_stlf_val
) |>
  pivot_longer(-date, names_to = "series", values_to = "load") |>
  ggplot(aes(x = date, y = load, color = series)) +
  geom_line() +
  scale_color_manual(values = c(observed = "lightgrey",
                                forecast = "blue")) +
  labs(x = NULL, y = "Daily avg load (Wh)", color = NULL,
       title = "Model 4: STL+ETS -- July 2010 validation") +
  theme_minimal()
```

### Validation Performance

STL+ETS achieved a validation MAPE of 14.49% and RMSE of 705.2 on the July 2010 hold-out. While this model handles both weekly and annual seasonality, it did not perform as well as TBATS and SARIMA on this dataset.

### Strengths and Limitations

STL+ETS is not our best model compared to TBATS and SARIMA.
However, it handles multiple seasonality well through
decomposition and is robust to outliers through the ETS
component.

------------------------------------------------------------------------

## Model 5: NNETAR

### Motivation

We decided to add NNETAR to our list because NNETAR feeds lagged values of the series into a single-hidden-layer feed-forward neural network, capturing non-linear relationships between past and future observations. We included it to test whether non-linear patterns in daily load.

### Specification

We applied nnetar() to the msts training object with set.seed(123) for reproducibility. The function automatically determined the number of input lags and hidden nodes to minimize in-sample error.

```{r model-5, eval=FALSE}
set.seed(123)
y_tr_nnetar   <- msts(train2010$load_day,
                      seasonal.periods = c(7, 365.25))
fit_nnetar    <- nnetar(y_tr_nnetar)
print(fit_nnetar)

fc_nnetar_val <- as.numeric(
  forecast(fit_nnetar, h = nrow(val2010))$mean)

mape_nnetar <- mape_pct(val2010$load_day, fc_nnetar_val)
rmse_nnetar <- rmse_val(val2010$load_day, fc_nnetar_val)

cat(sprintf("NNETAR -- Val MAPE: %.2f%%  |  Val RMSE: %.1f\n",
            mape_nnetar, rmse_nnetar))

model_scores <- rbind(model_scores, data.frame(
  Model       = "NNETAR",
  Predictors  = "Load only",
  Val_MAPE    = round(mape_nnetar, 2),
  Val_RMSE    = round(rmse_nnetar, 1),
  Kaggle_MAPE = "—"
))

data.frame(
  date     = val2010$date,
  observed = val2010$load_day,
  forecast = fc_nnetar_val
) |>
  pivot_longer(-date, names_to = "series", values_to = "load") |>
  ggplot(aes(x = date, y = load, color = series)) +
  geom_line() +
  scale_color_manual(values = c(observed = "lightgrey",
                                forecast = "blue")) +
  labs(x = NULL, y = "Daily avg load (Wh)", color = NULL,
       title = "Model 5: NNETAR -- July 2010 validation") +
  theme_minimal()
```

### Validation Performance

NNETAR achieved a validation MAPE of 14.95% and RMSE of 807.4 on the July 2010 hold-out. This is the weakest performance among all five models, indicating that the neural network overfit the training data and did not generalize well to the validation period.

### Strengths and Limitations

NNETAR is not our best model due to overfitting, where the
model performed well on training data but poorly on the
validation set. However, it does not require stationarity
assumptions and is flexible in capturing non-linear patterns
in the data.

------------------------------------------------------------------------

# Comparative Summary of the Five Models

## Performance Comparison Table

```{r comparison-table}
model_scores |>
  kable(
    col.names = c("Model", "Predictors", "Val MAPE (%)", "Val RMSE", "Kaggle MAPE"),
    align     = c("l", "l", "r", "r", "r"),
    caption   = "Validation performance of the top five models (July 2010 hold-out)."
  ) |>
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE)
```

[Insert a short paragraph summarizing the table. E.g., which model performed best on validation, and whether Kaggle scores were consistent with local MAPE.]

## Cross-Model Interpretation

-   **Seasonality capture:** TBATS explicitly models both weekly and annual seasonality. [Compare with Models 2–5 once finalized.]
-   **Responsiveness to short-term changes:** [Compare models' ability to track week-to-week variation in July.]
-   **Usefulness of weather covariates:** [If Models 3–5 incorporate temperature/humidity, comment on the gain.]
-   **Stability vs. flexibility:** [Simpler models may be more robust when validation data is short (31 days).]

## Final Model Selection

```{r final-model-refit, cache=TRUE}
# Retrain the chosen model on the full dataset (through 2011-06-30)
# and generate July 2011 forecasts.
# Submission CSVs are produced by the dedicated scripts in scripts/.

y_full_tbats <- msts(train_full$load_day, seasonal.periods = c(7, 365.25))

fit_final <- tbats(y_full_tbats)

fc_july2011 <- as.numeric(forecast(fit_final, h = 31)$mean)

tpl <- read.csv("data/submission_template.csv", stringsAsFactors = FALSE)
tpl$load <- round(fc_july2011, digits = 2)

head(tpl)
```

We selected **[Model name]** for the final submission because [brief justification — e.g., lowest validation MAPE, best Kaggle score, or best balance of accuracy and interpretability].

# Discussion

## Key Findings

[Summarize the main findings from the modeling exercise. E.g., which seasonal structure was most important, how large MAPE values were relative to the competition baseline, etc.]

## Role of Weather Variables

[Discuss whether temperature and/or humidity improved forecasting performance. If Models 3–5 include weather predictors, compare their MAPE to the load-only models.]

## Limitations

-   **Short validation window:** A single 31-day July hold-out may not fully characterize model performance across seasons.
-   **Single-meter data:** Results may not generalize to aggregate or grid-level load.
-   **External predictors:** Only daily temperature and humidity averages were used; hour-by-hour or maximum temperature values might improve performance.

## Future Improvements

-   Dynamic harmonic regression with ARIMA errors (Fourier terms for annual seasonality within an ARIMA framework).
-   Neural network models (NNETAR) or gradient boosting with calendar and weather features.
-   Ensembling multiple models (weighted average of ARIMA and TBATS predictions).

# Conclusion

This project developed and compared five time series forecasting models to predict daily residential electricity demand for July 2011. We began with TBATS to jointly capture weekly and annual seasonality, and further explored [brief description of Models 2–5 once finalized]. Across the five models evaluated, [Model X] achieved the lowest validation MAPE of [X]%, and was selected for the final Kaggle submission. The results highlight the importance of explicitly modeling dual seasonality in daily load data.

# References and Acknowledgments

## References

-   Hyndman, R. J., & Athanasopoulos, G. (2021). *Forecasting: Principles and Practice* (3rd ed.). OTexts. <https://otexts.com/fpp3/>
-   R Core Team (2024). *R: A Language and Environment for Statistical Computing*. R Foundation for Statistical Computing.
-   Hyndman, R., et al. (2024). *forecast: Forecasting Functions for Time Series and Linear Models*. R package version 8.x. <https://pkg.robjhyndman.com/forecast/>
-   [Additional course materials as appropriate]

## AI Use Disclosure

> LLMs (e.g., Claude) were used to assist with code structuring, report organization, and language polishing. All modeling decisions, parameter choices, interpretation of results, and final submissions were completed and verified by us.
